Computer Assisted Verbal Autopsy: Comparing Large Language Models to Physicians for Assigning Causes to 6939 Deaths in Sierra Leone from 2019-2022

### 2025-06-23 manuscript/manuscript.tex
Editing methods section

\abstract{\textbf{Background:} Verbal Autopsies (VAs) collect data on deaths and their causes outside of traditional hospital settings to provide representative Causes of Death (CODs). Current computer models for COD assignment in VAs perform similar to physicians at the population level, but poorly at the individual level, due to focuses on questionnaire data and neglecting free text from narratives. Recently, a large language model called ChatGPT-4 demonstrated human-level performance on professional and academic exams. ChatGPT-4 shows promise in assigning CODs similar to physicians, but has yet been examined for assigning CODs using VA narratives.

\textbf{Methods:} 6939 VA records in Sierra Leone from 2019 to 2022 were used to compare four computer models, GPT-3.5, GPT-4, InterVA-5, and InSilicoVA, to physician COD assignment at population and individual levels. Narratives were used for GPT-3.5/4, while questionnaires were used for InterVA-5/InSilicoVA. COD assignments were grouped into general COD categories consisting of 19, 10, and 7 categories for adult, child, and neonatal age groups. Cause Specific Mortality Fraction (CSMF) accuracy and Partial Corrected Concordance (PCCC) were used to compare models to physicians at population and individual levels respectively. CSMF and PCCC were evaluated overall and by COD, age group, and age.

\textbf{Results:} GPT-4 had the best performance overall (0.61 PCCC), followed by GPT-3.5 (0.56 PCCC), and InSilicoVA/InterVA-5 (0.44 PCCC). GPT-4 had the best performance for adult and neonatal records (0.64 and 0.58 PCCC), with GPT-3.5 for child records (0.54 PCCC). All models' performances trended upwards from 1 month to 14 years ($\sim$0.1-0.75 PCCC) and downwards from 15-69 years ($\sim$0.7-0.35) of age. GPT-4, GPT-3.5, and InSilicoVA had the highest performances for 17, 9, and 4 of all 30 CODs respectively. At the population level, all models had CSMF accuracies between 0.74-0.79.

\textbf{Conclusion:} All models performed similarly at the population level, while GPT-3.5/4 and InSilicoVA performed better for several CODs at the individual level. GPT models made improvements over InSilicoVA and InterVA-5 models. Our research lays the foundation for future work in computer assisted VA, where physicians utilize alternative COD assignments from computer models to help reduce ill-defined codes and physician disagreement.
}

@@

In 2019, 41 million people died prematurely from noncommunicable diseases every year, accounting for 74\% of all deaths globally \citep{worldhealthorganizationNonCommunicableDiseases2019}. Most of these deaths are preventable, but require adequate resource allocation, guided by evidence, to implement effective interventions and policies that target populations at risk \citep{benzigerGlobalBurdenDisease2016}. Thus, reliable counts and diagnoses of deaths enable decision makers to identify populations at risk to save lives and reduce premature deaths worldwide \citep{lawnMillionNeonatalDeaths2010,lassiCommunityBasedIntervention2015,liuExcessMortalityPersons2017,ewigCommunityacquiredPneumoniaEmergency2011}. However, many low-income countries do not have data on deaths or have registered less than half of the deaths in their country, with an even fewer 8\% of these registered deaths having a Cause of Death (COD) recorded \citep{worldhealthorganizationSCOREHealthData2021}. To fill this gap in death registrations, an alternative method known as Verbal Autopsy (VA) is used to collect data on deaths and determine their likely causes at scale \citep{desavignyIntegratingCommunitybasedVerbal2017,thomasVerbalAutopsyHealth2018,rampatigeSystematicReviewStatistics2014}, outside of traditional healthcare facilities where over half of deaths occur at home \citep{adairWhoDiesWhere2021}.

VA involves two major components: survey and COD assignment \citep{worldhealthorganizationVerbalAutopsyStandards2023,chandramohanEstimatingCausesDeath2021,worldhealthorganizationVerbalAutopsyStandards2007}. In the survey component, trained lay surveyors interview those familiar with the deceased (e.g. living spouse, children, family, friends) to gather information using standardized questionnaires and open narratives. In the COD assignment component, physicians evaluate information available from the questionnaires and open narratives to assign probable CODs. This component has been criticized to be difficult to reproduce due to reliance on physician assignment \citep{gomesNationwideMortalityStudies2017,jhaProspectiveStudyOne2005,mccormickProbabilisticCauseofDeathAssignment2016b,morrisFactorsAssociatedPhysician2010,solemanVerbalAutopsyCurrent2006}. As an alternative to physician assignment, computer models, such as InterVA \citep{byassIntegratedApproachProcessing2019} and InSilicoVA \citep{mccormickProbabilisticCauseofDeathAssignment2016b}, have been studied to automatically assign CODs with performances close to physicians at the population level, but poor performances at the individual level \citep{jhaAutomatedPhysicianAssignment2019,leitaoComparisonPhysiciancertifiedVerbal2014,desaiPerformanceFourComputercoded2014,tungaVerbalAutopsyModels2021,otiVerbalAutopsyInterpretation2010}. These computer models often utilize data from the structured questionnaire, but often omit the free-text open narrative, which misses latent information, such as chronology or health-seeking behaviors, that may potentially help models perform better than using the questionnaire alone \citep{jebleeAutomaticallyDeterminingCause2019,blancoExtractingCauseDeath2021,kingQualityDiagnosticValue2016}.

Recently, Large Language Models (LLM), leveraging massive datasets and deep learning approaches, have made advances in performing a variety of Natural Language Processing (NLP) tasks using free-text, such as question answering, code generation, and even medical diagnosis \citep{changSurveyEvaluationLarge2023,lundChattingChatGPTHow2023,svyatkovskiyIntelliCodeComposeCode2020,hauptAIGeneratedMedicalAdvice2023}. In 2022, a widely-available LLM called ChatGPT was released by OpenAI with capabilities of answering natural language text inquiries using training data up to September 2021. ChatGPT-3 was based on several Generative Pre-trained Transformer (GPT) models between 2018 to 2020, namely GPT-1 to GPT-3, which had notable differences in training data sizes of 5 gigabytes to 45 terabytes from web sources that resulted in 117 million to 175 billion parameter models \citep{wuBriefOverviewChatGPT2023}. In March 2023, ChatGPT-4 was released with human-level performance on various professional and academic exams and benchmarks that outperformed ChatGPT-3 \citep{openaiGPT4TechnicalReport2023}. Given the limited usage of free-text open narratives in computer models for determining CODs, and recent advances in LLMs that leverage natural language text prompts, a study was conducted for Sierra Leone deaths from VA in 2019 to 2022 to compare four models, GPT-3.5, GPT-4, InterVA-5, and InSilicoVA, to physicians for determining CODs.

@@

This study details the methods used to compare the COD assignment from four models, GPT-3.5, GPT-4, InterVA-5, and InSilicoVA, to physicians as seen in Figure \ref{fig-methods-brief}. The initial VA data was filtered for physician agreed records as described in Section \ref{methods-data}. Section \ref{methods-models} describes the input and output of the four models for COD assignment, while section \ref{methods-eval} details the performance evaluation of the models relative to physicians using population and individual level metrics. See Appendix \ref{app-methods} for additional details on the methods used in this study.

@@

    \botrule
    \end{tabular}
    \footnotetext[1]{Other infections and severe systemic/localized infections were generalized into infections. Appendix, hernia, intestinal and Peptic ulcer/gastroesophageal causes did not have comparable CGHR-10 codes and were omitted.}
    \footnotetext[2]{Percentage of $\sim$76 Thousand (K) total deaths \citep{ansumanaReportCausesDeath2023}. Numbers are rounded.}
    \footnotetext[3]{Adult, child, and neonate mean PCCC and summed n records if available.}
\end{table}

>>>>

\abstract{
\textbf{Background:} Verbal autopsies (VAs) collect information on deaths occurring outside traditional healthcare settings to estimate representative causes of death (CODs). Current computer models assign CODs at population-level accuracy comparable to physicians, but perform poorly at the individual level, largely due to reliance on structured questionnaire data and neglect of narrative free text. Recently, the large language model ChatGPT-4 demonstrated human-level performance on professional and academic benchmarks. While ChatGPT-4 shows promise in COD assignment, its application to VA narratives has not yet been evaluated.

\textbf{Methods:} We analyzed 6,939 VA records from Sierra Leone (2019–2022) to compare four models, GPT-3.5, GPT-4, InterVA-5, and InSilicoVA, against physician-assigned CODs at both population and individual levels. GPT models used narratives, whereas InterVA-5 and InSilicoVA relied on questionnaires. CODs were grouped into 19, 10, and 7 categories for adult, child, and neonatal deaths. Cause Specific Mortality Fraction (CSMF) accuracy and Partial Chance Corrected Concordance (PCCC) were used to assess population and individual level agreement with physician coding respectively, stratified by age and COD.

\textbf{Results:} GPT-4 outperformed all models overall (PCCC=0.61), followed by GPT-3.5 (0.56) and InSilicoVA/InterVA-5 (0.44). GPT-4 achieved the highest PCCC for adult and neonatal deaths (0.64/0.58), with GPT-3.5 for child deaths (0.54). Across ages, model performance increased from 1 month to 14 years ($\sim$0.10–0.75 PCCC) and declined from 15 to 69 years ($\sim$0.70–0.35). GPT-4, GPT-3.5, and InSilicoVA achieved the highest PCCC in 17, 9, and 4 of the 30 CODs, respectively. At the population level, all models achieved comparable CSMF accuracies (0.74–0.79).

\textbf{Conclusion:} All models performed similarly at the population level, but GPT models and InSilicoVA showed greater accuracy for specific CODs at the individual level. GPT models demonstrated improvements over InterVA-5 and InSilicoVA models. This study provides foundational evidence for integrating large language models into computer assisted VA to support physicians, reducing ill-defined codes and improving agreement in COD assignment.
}

@@

In 2019, 41 million people died prematurely from noncommunicable diseases every year, accounting for 74\% of all deaths globally \citep{worldhealthorganizationNonCommunicableDiseases2019}. While most of these deaths are preventable, effective intervention requires evidence-based resource allocation targeting high-risk populations \citep{benzigerGlobalBurdenDisease2016}. Reliable mortality counts and accurate Cause of Death (COD) data are therefore essential for guiding public health policy and reducing premature mortality \citep{lawnMillionNeonatalDeaths2010,lassiCommunityBasedIntervention2015,liuExcessMortalityPersons2017,ewigCommunityacquiredPneumoniaEmergency2011}. However, in many low-income countries, civil registration and vital statistics systems remain incomplete. Fewer than half of all deaths are registered, and among these, only 8\% have an assigned COD \citep{worldhealthorganizationSCOREHealthData2021}. To address this gap, Verbal Autopsy (VA) has been employed as a scalable method for collecting mortality data and assigning likely CODs, particularly for deaths that occur outside of healthcare facilities, which account for more than half of all deaths in these settings \citep{desavignyIntegratingCommunitybasedVerbal2017,thomasVerbalAutopsyHealth2018,rampatigeSystematicReviewStatistics2014,adairWhoDiesWhere2021}.

VA involves two major components: survey and COD assignment \citep{worldhealthorganizationVerbalAutopsyStandards2023,chandramohanEstimatingCausesDeath2021,worldhealthorganizationVerbalAutopsyStandards2007}. In the survey component, trained interviewers use structured questionnaires and open narrative prompts to gather data from relatives or close contacts of the deceased. In the COD assignment component, physicians review these data to determine the most likely COD. However, reliance on physician assignment has been criticized for limited reproducibility and subjectivity \citep{gomesNationwideMortalityStudies2017,jhaProspectiveStudyOne2005,mccormickProbabilisticCauseofDeathAssignment2016b,morrisFactorsAssociatedPhysician2010,solemanVerbalAutopsyCurrent2006}. To overcome these limitations, automated Computer Coded Verbal Autopsy (CCVA) methods such as InterVA \citep{byassIntegratedApproachProcessing2019} and InSilicoVA \citep{mccormickProbabilisticCauseofDeathAssignment2016b} have been developed. These models offer scalable and reproducible alternatives and have demonstrated comparable performance to physicians at the population level. However, their performance at the individual level remains limited \citep{jhaAutomatedPhysicianAssignment2019,leitaoComparisonPhysiciancertifiedVerbal2014,desaiPerformanceFourComputercoded2014,tungaVerbalAutopsyModels2021,otiVerbalAutopsyInterpretation2010}, while their reliance on structured questionnaire data often omits open narrative text, which can contain additional contextual and chronological information that may improve diagnostic accuracy \citep{jebleeAutomaticallyDeterminingCause2019,blancoExtractingCauseDeath2021,kingQualityDiagnosticValue2016}.

Recent advances in large language models (LLMs), trained on vast textual datasets using deep learning methods, have significantly improved natural language processing (NLP) capabilities. These include tasks such as question answering, code generation, and medical reasoning based on free text \citep{changSurveyEvaluationLarge2023,lundChattingChatGPTHow2023,svyatkovskiyIntelliCodeComposeCode2020,hauptAIGeneratedMedicalAdvice2023}. ChatGPT, developed by OpenAI and released in 2022, is a widely accessible LLM capable of generating human-like responses to natural language queries. Earlier versions (GPT-1 to GPT-3) scaled from 117 million to 175 billion parameters and were trained on data ranging from 5 GB to 45 TB \citep{wuBriefOverviewChatGPT2023}. In 2023, ChatGPT-4 was introduced, achieving human-level performance on a range of academic and professional benchmarks \citep{openaiGPT4TechnicalReport2023}. Given the underutilization of narrative free text in VA analysis and the capabilities of LLMs in processing such data, we conducted a study using VA records from Sierra Leone (2019–2022) to compare four models, GPT-3.5, GPT-4, InterVA-5, and InSilicoVA, against physician-assigned CODs. This work aims to evaluate the potential of LLMs in enhancing COD assignment from narrative data in low-resource settings.

@@

This study outlines the methodology used to compare cause of death (COD) assignments from four models, GPT-3.5, GPT-4, InterVA-5, and InSilicoVA, with physician-determined CODs, as summarized in Figure \ref{fig-methods-brief}. The dataset was first filtered to include only records with physician agreement, as described in Section \ref{methods-data}. Section \ref{methods-models} details the input formats and output structures of the four models. Section~\ref{methods-eval} presents the evaluation framework, which compares model outputs to physician-assigned CODs using both population-level and individual-level performance metrics. Additional methodological details are provided in Appendix \ref{app-methods}.

@@

    \botrule
    \end{tabular}
    \footnotetext[1]{Other infections and severe systemic/localized infections were generalized into infections. Appendix, hernia, intestinal and Peptic ulcer/gastroesophageal causes did not have comparable CGHR-10 codes and were omitted from the top ten.}
    \footnotetext[2]{Percentage of $\sim$76 Thousand (K) total deaths \citep{ansumanaReportCausesDeath2023}. Numbers are rounded.}
    \footnotetext[3]{Adult, child, and neonate mean PCCC and summed n records if available.}
\end{table}
